---
title: "Integrating single-cell samples with Conos"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integrating single-cell samples with Conos}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4.5)
```

**Conos** (Clustering On Network of Samples) jointly analyses a *collection* of single-cell samples.
Rather than merging cells into one matrix, it aligns each pair of samples in a shared reduced space,
stitches the per-sample neighbourhoods into one weighted graph, and runs clustering, embedding and
differential expression on that joint graph. This keeps each sample's own structure intact while still
letting corresponding cell populations connect across samples.

This vignette walks through the minimal workflow on a small bundled example. For larger, fully worked
tutorials (including the choice of alignment space, data sources, and multimodal data) see the rendered
notebooks at <https://github.com/kharchenkolab/conos> (the `doc/` directory of the source repository).

## Input

Conos takes a named list of already-preprocessed per-sample objects (`Pagoda2` or `Seurat`). The package
ships a tiny two-sample panel so the vignette builds quickly:

```{r}
library(conos)
data("small_panel.preprocessed", package = "conos")
names(small_panel.preprocessed)
```

Each element is one sample with its own PCA reduction; to prepare your own data, process each sample with
pagoda2 (e.g. `Pagoda2$new()` followed by `runVariance()` and a PCA reduction) or Seurat before passing it
to Conos.

## Building the joint graph

```{r}
con <- Conos$new(small_panel.preprocessed, n.cores = 1)
```

`runGraph()` aligns every pair of samples and builds the joint graph. `space = "PCA"` uses reciprocal PCA
(the recommended default); `k`/`k.self` set the inter- and within-sample neighbourhood sizes.

```{r}
con$runGraph(k = 15, k.self = 5, space = "PCA", ncomps = 20, n.odgenes = 1000, verbose = FALSE)
```

## Clustering and embedding

```{r}
con$runClustering(verbose = FALSE)                 # Leiden communities on the joint graph
con$runEmbedding(method = "UMAP", verbose = FALSE)  # joint 2D embedding
```

Colour the joint graph embedding by the detected clusters:

```{r}
con$plotGraph(clustering = "leiden")
```

`plotPanel()` shows the same joint clustering laid over each sample separately, which is useful for
checking that clusters are shared rather than sample-specific:

```{r}
con$plotPanel(clustering = "leiden", use.common.embedding = FALSE)
```

## Markers

`runMarkers()` finds genes differentially expressed between clusters, and `plotMarkerDotPlot()` shows the
top markers per cluster (dot size = fraction of cells expressing; colour = scaled mean expression):

```{r}
con$plotMarkerDotPlot(n.genes.per.group = 3, z.threshold = 0.5)
```

## Transferring labels

If one sample (or a subset of cells) is annotated, `propagateLabels()` spreads those labels over the joint
graph and reports a per-cell confidence:

```{r}
groups <- con$clusters$leiden$groups
known  <- groups[sample(length(groups), round(0.7 * length(groups)))]  # pretend 70% are annotated
res <- con$propagateLabels(labels = known, verbose = FALSE)
str(res$labels)        # transferred label per cell
summary(res$uncertainty)
```

## Where to go next

* `?Conos` — the full set of methods. The preferred verbs are `runGraph()`, `runClustering()`,
  `runEmbedding()` and `runMarkers()` (the older `buildGraph()`/`findCommunities()`/`embedGraph()`/
  `getDifferentialGenes()` names still work but are deprecated).
* `planIntegration()` — inspect, before building, which modalities/features the samples actually share.
* The rendered notebooks at <https://github.com/kharchenkolab/conos> — a fuller novice walkthrough and an
  advanced tutorial (alignment spaces, data sources, multimodal facets, label transfer).
