---
title: "Convert Conos Object to ScanPy"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{"Convert Conos Object to ScanPy"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

First load the conos library, and the R package `conosPanel` for the example data `panel`:

```{r, message=FALSE, warning=FALSE}
library(conos)
panel <- conosPanel::panel
```

Next, use pagoda2 for pre-processing:

```{r, message=FALSE, warning=FALSE}
library(pagoda2)
panel.preprocessed <- lapply(panel, basicP2proc, n.cores=2, min.cells.per.gene=0, n.odgenes=2e3, 
                             get.largevis=FALSE, make.geneknn=FALSE)
```

Now align the datasets:

```{r, message=FALSE, warning=FALSE}
con <- Conos$new(panel.preprocessed, n.cores=2)
con$buildGraph(k=15, k.self=5, space='PCA', ncomps=30)
```


Next find the clusters, and create an embedding:

```{r, message=FALSE, warning=FALSE}
con$findCommunities()
con$embedGraph(method="UMAP")
```

Now prepare the metadata (which can be any type of clustering of all the cells):

```{r, message=FALSE, warning=FALSE}
metadata <- data.frame(Cluster=con$clusters$leiden$groups)
```

Save data (set `exchange_dir` to your path):


```{r, message=FALSE, warning=FALSE}
## use current directory
exchange_dir <- "."
hdf5file = "example.h5"
saveConosForScanPy(con, output.path=exchange_dir, hdf5_filename=hdf5file, verbose=TRUE)
```

Users can then access the data saved to the HDF5 file, e.g. to access metadata, run:


```{r, message=FALSE, warning=FALSE}
library(rhdf5)
metadata = h5read(paste0(exchange_dir, "/example.h5"), 'metadata/metadata.df')
head(metadata, 4)
```

All possible fields included in the output HDF5 file are:

* `raw_count_matrix`: the sparse `dgCMatrix` of raw counts 
	* `raw_count_matrix/data`: the matrix entries
	* `raw_count_matrix/shape`: the matrix dimensions
	* `raw_count_matrix/indices`: 0-based vector of non-zero matrix entries
	* `raw_count_matrix/indptr`: vector of pointers, one for each column (or row), to the initial (zero-based) index of elements in the column (or row)
* `metadata/metadata.df`: the `data.frame` of metadata values
* `genes/genes.df`: the `data.frame` of genes
* `count_matrix`: the sparse `dgCMatrix` of normalized counts, if `cm.norm=TRUE`
	* `count_matrix/data`: the matrix entries
	* `count_matrix/shape`: the matrix dimensions
	* `count_matrix/indices`: 0-based vector of non-zero matrix entries
	* `count_matrix/indptr`: vector of pointers, one for each column (or row), to the initial (zero-based) index of elements in the column (or row)
* `embedding/embedding.df`:  the `data.frame` of the conos embedding, if `embedding=TRUE`
* `pseudopca/pseudopca.df`:  the `data.frame` of an emulated PCA by embedding the graph to a space with `n.dims` dimensions and save it as a pseudoPCA, if `pseudopca=TRUE`
* `pca/pca.df`: the `data.frame` of the PCA of all the samples (not batch corrected), if `pca=TRUE`
* `graph_connectivities`: the sparse `dgCMatrix` of graph connectivites, if `alignment.graph=TRUE`
	* `graph_connectivities/data`: the matrix entries
	* `graph_connectivities/shape`: the matrix dimensions
	* `graph_connectivities/indices`: 0-based vector of non-zero matrix entries
	* `graph_connectivities/indptr`: vector of pointers, one for each column (or row), to the initial (zero-based) index of elements in the column (or row)
* `graph_distances`: the sparse `dgCMatrix` of graph distances, if `alignment.graph=TRUE`
	* `graph_distances/data`: the matrix entries
	* `graph_distances/shape`: the matrix dimensions
	* `graph_distances/indices`: 0-based vector of non-zero matrix entries
	* `graph_distances/indptr`: vector of pointers, one for each column (or row), to the initial (zero-based) index of elements in the column (or row)



In order to read in the `dcGMatrix` again, simply use the `Matrix` package as follows:


```{r, message=FALSE, warning=FALSE}
library(rhdf5)
library(Matrix)
rawcountMat = h5read(paste0(exchange_dir, "/example.h5"), 'raw_count_matrix')
raw_count_matrix = sparseMatrix(x = as.numeric(rawcountMat$data),  
    dims = as.numeric(c(rawcountMat$shape[1], rawcountMat$shape[2])), 
    p = as.numeric(rawcountMat$indptr), 
    i = rawcountMat$indices, index1=FALSE)
```

**Note:** Please set `index1=FALSE` as the index vectors are 0-based. For more details, see the documentation for [Matrix::sparseMatrix()](https://www.rdocumentation.org/packages/Matrix/versions/1.2-18/topics/sparseMatrix)

