---
title: "Conos walkthrough: jointly analysing a panel of samples"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    self-contained: true
  ipynb: default
jupyter: ir44
---

**Conos** (*Clustering On Network of Samples*) analyses a *collection* of single-cell samples **together**.
Rather than merging all the cells into one matrix, it aligns each pair of samples in a shared space and
stitches their neighbourhoods into a single weighted graph. On that joint graph it then runs clustering,
embedding, and differential expression — so cells of the same type connect *across* samples while each
sample keeps its own structure.

We run the whole standard workflow on a panel of four bone-marrow (BM) and cord-blood (CB) samples. When
you need to control *how* the samples are aligned (alignment spaces, strength, memory, multimodal data),
see the [advanced tutorial](conos-advanced.ipynb).

## The workflow at a glance

The entire standard analysis is just a handful of calls on the Conos object. The rest of this walkthrough
runs through each line, explains the parameters, and shows what to look for in the output:

```r
# one Pagoda2 object per sample: normalisation + PCA (+ a within-sample clustering)
samples <- lapply(panel, function(counts) Pagoda2$from(counts)$run())

con <- Conos$new(samples)
con$runGraph()           # align every pair of samples into one joint graph
con$runClustering()      # joint clusters (Leiden communities)
con$runEmbedding()       # joint 2-D embedding (UMAP)

con$plotGraph()                          # joint embedding: colour by cluster / sample / gene
con$plotPanel()                          # the same, faceted per sample
con$plotMarkerDotPlot()                  # top marker genes per cluster
con$propagateLabels(labels = cellannot)  # transfer annotations from one sample to the rest
```

## Contents

- [The workflow at a glance](#the-workflow-at-a-glance)
- [Loading the data](#loading-the-data)
- [Pre-processing each sample](#pre-processing-each-sample)
- [Constructing the Conos object](#constructing-the-conos-object)
- [Building the joint graph](#building-the-joint-graph)
- [Joint clustering](#joint-clustering)
- [The joint embedding](#the-joint-embedding)
- [Shared clusters vs per-sample clusters](#shared-clusters-vs-per-sample-clusters)
- [Are the samples well mixed?](#are-the-samples-well-mixed)
- [Cluster composition across samples](#cluster-composition-across-samples)
- [Looking at a gene](#looking-at-a-gene)
- [Marker genes](#marker-genes)
- [Transferring annotations between samples](#transferring-annotations-between-samples)
- [Markers of the transferred cell types](#markers-of-the-transferred-cell-types)

## Loading the data

A Conos panel is a **named list of per-sample objects** — one per sample, each processed with
[pagoda2](https://github.com/kharchenkolab/pagoda2) (or Seurat). You build one object per sample from its
own count data; `Pagoda2$from()` reads the formats you are most likely to have:

```{r}
#| eval: false
library(pagoda2)

# a 10x CellRanger output folder (matrix.mtx / barcodes / features):
s1 <- Pagoda2$from10x("sample1/outs/filtered_feature_bc_matrix")

# a 10x HDF5 file:
s2 <- Pagoda2$from10xH5("sample2_filtered_feature_bc_matrix.h5")

# a counts matrix (genes x cells) you already have in memory:
s3 <- Pagoda2$from(my_counts_matrix)
```

For a runnable example we use a ready-made panel from the `conosPanel` package
(`install.packages("conosPanel", repos = "https://kharchenkolab.github.io/drat/")`). It is a named list of
four sparse count matrices (genes × cells) — two bone-marrow and two cord-blood samples:

```{r}
#| message: false
#| warning: false
library(conos)
library(pagoda2)

panel <- conosPanel::panel
str(panel, 1)
```

**Cell names must be unique across the whole panel** — no two cells, even in different samples, may share a
name. Here each barcode is already prefixed with its sample id, so there are no collisions:

```{r}
head(colnames(panel[[1]]), 3)
any(duplicated(unlist(lapply(panel, colnames))))   # FALSE = all unique
```

To keep the tutorial fast we down-sample each matrix to a thousand cells; in a real analysis you would use
them all (and would first filter out low-count / dying cells as part of standard QC).

```{r}
set.seed(1)
panel <- lapply(panel, function(m) m[, sample(ncol(m), 1000)])
```

## Pre-processing each sample

Each sample is turned into a `Pagoda2` object and processed **on its own**: normalisation, a PCA reduction
(which Conos will align on), a within-sample kNN graph, and a within-sample clustering. A single `run()`
call does all of it, so the whole panel is one short `lapply`:

```{r}
#| message: false
#| warning: false
samples <- lapply(panel, function(counts) {
  Pagoda2$from(counts, verbose = FALSE)$run(verbose = FALSE)
})
```

We now have a named list of `Pagoda2` objects — the starting point for Conos:

```{r}
names(samples)
```

(If you prefer Seurat, pre-process with `basicSeuratProc()` instead and pass the resulting list straight to
`Conos$new()`; everything below is identical.)

## Constructing the Conos object

`Conos$new()` simply collects the per-sample objects — nothing is computed yet. `n.cores` controls the
parallelism of the steps that follow (graph construction in particular benefits from more cores):

```{r}
con <- Conos$new(samples, n.cores = 1)
```

## Building the joint graph

`runGraph()` is the heart of Conos. For **every pair** of samples it projects the two into a shared space,
finds mutual nearest neighbours between them, and adds within-sample neighbours so that every cell is
connected — producing one joint graph spanning the whole panel.

The `space` argument chooses how each pair is matched. `"PCA"` (reciprocal PCA, the default used here) is
fast and robust and is the right choice for most panels; `"CCA"` helps for very heterogeneous panels;
`"CPCA"` and `"genes"` are also available. The [advanced tutorial](conos-advanced.ipynb) compares them.

```{r}
#| message: false
con$runGraph(k = 15, k.self = 5, space = "PCA", ncomps = 30, n.odgenes = 2000, verbose = FALSE)
```

## Joint clustering

Now detect cell clusters as **communities** on the joint graph (Leiden by default). Because the graph spans
all samples, each cluster is a population defined *consistently* across the whole panel:

```{r}
#| message: false
con$runClustering(verbose = FALSE)
```

## The joint embedding

Compute a 2-D embedding of the joint graph (UMAP) so we can see it, then colour it by the clusters:

```{r}
#| message: false
con$runEmbedding(method = "UMAP", min.visited.verts = 100, verbose = FALSE)
```

```{r}
#| fig-width: 6
#| fig-height: 5
con$plotGraph(clustering = "leiden")
```

Each blob is a cluster; neighbouring blobs are transcriptionally similar populations.

## Shared clusters vs per-sample clusters

Why bother with a joint graph at all? Each sample was *also* clustered on its own during pre-processing.
Laid over the joint embedding, those **independent** clusterings use a different, unrelated set of labels in
every sample — cluster "3" in one sample has nothing to do with cluster "3" in another:

```{r}
#| fig-width: 8
#| fig-height: 6
con$plotPanel(clustering = "leiden", use.local.clusters = TRUE, use.common.embedding = TRUE, font.size = 3)
```

The **joint** clustering fixes exactly this. The same colours now mean the same population in every sample,
so a cluster present in several samples is immediately recognisable as one shared cell type:

```{r}
#| fig-width: 8
#| fig-height: 6
con$plotPanel(clustering = "leiden", use.common.embedding = TRUE)
```

## Are the samples well mixed?

*The* question for any integration: do the samples **overlap** in the joint embedding (good — the same cell
types are aligned across samples) or sit apart (under-correction, or genuinely distinct populations)? Colour
the *same* embedding by **sample**:

```{r}
#| fig-width: 6.5
#| fig-height: 5
con$plotGraph(color.by = "sample", mark.groups = FALSE, alpha = 0.1, show.legend = TRUE)
```

Cells from all four samples are interspersed within each region rather than forming sample-specific islands,
so the populations are well aligned.

## Cluster composition across samples

`plotClusterBarplots()` shows what fraction of each cluster comes from each sample — handy for spotting
populations that are sample- or condition-specific (here, differences between bone marrow and cord blood):

```{r}
#| fig-width: 7
#| fig-height: 4
plotClusterBarplots(con, clustering = "leiden", show.entropy = FALSE)
```

A bar split evenly across samples is a shared population; a bar dominated by one or two samples flags a
condition-specific one.

## Looking at a gene

You can paint any gene onto the joint embedding to see where it is expressed — here *CD3E*, a pan-T-cell
marker:

```{r}
#| fig-width: 6
#| fig-height: 5
con$plotGraph(gene = "CD3E")
```

The same gene, shown per sample on the common embedding, confirms the pattern is consistent across the
panel rather than driven by one sample:

```{r}
#| fig-width: 8
#| fig-height: 6
con$plotPanel(gene = "CD3E", use.common.embedding = TRUE)
```

## Marker genes

`runMarkers()` finds genes differentially expressed between clusters, and `plotMarkerDotPlot()` shows the
top few **specific** markers per cluster — dot **size** is the fraction of cells expressing the gene, dot
**colour** is the scaled mean expression:

```{r}
#| fig-width: 9
#| fig-height: 6
con$plotMarkerDotPlot(n.genes.per.group = 3)
```

The diagonal pattern (each cluster lit up by its own genes) is what to look for, and it lets you read off
cell identities — e.g. the T-cell (*CD3D/CD3E/IL7R*), B-cell (*MS4A1/CD79A*), monocyte (*FCN1/CSTA*) and
NK (*KLRF1/GZMB*) blocks.

## Transferring annotations between samples

A very common task: carry cell-type labels from one **annotated** sample onto the rest of the panel. We load
a real annotation of the first bone-marrow sample (shipped with conos) — a named factor of cell types:

```{r}
cellannot <- read.table(file.path(find.package("conos"), "extdata", "cellannot.txt"),
                        header = FALSE, sep = "\t")
cellannot <- setNames(cellannot[, 2], cellannot[, 1])
head(cellannot, 3)
```

Per sample, the labels exist for only one sample; the other three are unlabelled (shown faded):

```{r}
#| fig-width: 8
#| fig-height: 6
con$plotPanel(groups = cellannot, use.common.embedding = TRUE)
```

`propagateLabels()` spreads those labels over the joint graph, returning a transferred label **and** an
uncertainty score for every cell:

```{r}
#| message: false
transfer <- con$propagateLabels(labels = cellannot, verbose = FALSE)
```

On the joint embedding, the cell types now cover all the cells — and line up with the clusters we found
independently:

```{r}
#| fig-width: 6.5
#| fig-height: 5
con$plotGraph(groups = transfer$labels)
```

Faceting the transferred labels **by sample** shows the same cell types recovered consistently in every
sample, including the three that started out unlabelled:

```{r}
#| fig-width: 8
#| fig-height: 6
con$plotPanel(groups = transfer$labels, use.common.embedding = TRUE, show.legend = FALSE)
```

`transfer$uncertainty` is a 0–1 score per cell (low = confident). Painting it per sample shows *where* the
transfer is trustworthy — uncertainty rises at the boundaries between populations and in rare types:

```{r}
#| fig-width: 8
#| fig-height: 6
con$plotPanel(colors = transfer$uncertainty, use.common.embedding = TRUE, show.legend = TRUE)
```

## Markers of the transferred cell types

Because the transferred labels are *named* cell types, marker detection on them is directly interpretable.
`runMarkers()` returns, per cell type, a table of differentially expressed genes with effect size (`M`),
significance (`Z`), and — with `append.auc = TRUE` — discrimination and specificity metrics:

```{r}
#| message: false
de <- con$runMarkers(groups = transfer$labels, append.auc = TRUE, verbose = FALSE)
head(de[["B cells"]][order(-de[["B cells"]]$AUC), c("Gene", "M", "Z", "AUC", "Specificity", "Precision")])
```

The top B-cell markers are exactly the expected ones (*CD79A*, *MS4A1*, *IGHD*, MHC-II genes). A dot plot
across all the transferred types reads as a clean marker map of the panel:

```{r}
#| fig-width: 10
#| fig-height: 6
con$plotMarkerDotPlot(groups = transfer$labels, n.genes.per.group = 3)
```

That is the whole standard workflow previewed at the top, now seen end to end. To control *how* the
samples are aligned — alignment space, strength, memory, and multimodal data — continue to the
[advanced tutorial](conos-advanced.ipynb).
