---
title: "Conos walkthrough: integrating a panel of samples"
format:
  html:
    toc: true
    toc-depth: 3
    self-contained: true
  ipynb: default
jupyter: ir44
---

**Conos** (Clustering On Network of Samples) jointly analyses a *collection* of single-cell samples. Instead
of merging all cells into one matrix (which can bury real per-sample structure and amplify batch effects),
Conos aligns each *pair* of samples in a shared reduced space, stitches their neighbourhoods into one
weighted graph, and runs clustering, embedding, and differential expression on that joint graph. Cells of
the same type connect across samples, while each sample keeps its own structure.

This walkthrough takes the standard, minimal workflow end to end on a small bone-marrow / cord-blood panel.
It is deliberately bare — see the [advanced tutorial](conos-advanced.ipynb) for alignment-space choices,
memory controls, and label-transfer options.

## Setup

We use the example panel from the `conosPanel` package and preprocess each sample with `pagoda2`.

```{r}
#| message: false
#| warning: false
library(conos)
library(pagoda2)
library(conosPanel)

panel <- conosPanel::panel        # a named list of 4 sparse count matrices (genes x cells)
str(panel, 1)
```

To keep this tutorial quick to run we subsample each sample to ~1000 cells; in a real analysis you would
use all of them.

```{r}
set.seed(1)
panel <- lapply(panel, function(m) m[, sample(ncol(m), min(1000, ncol(m)))])
```

A Conos panel is just a named list of count matrices — one per sample. Cell names must be unique across the
whole panel (so cells from different samples never collide):

```{r}
any(duplicated(unlist(lapply(panel, colnames))))
```

## Preprocess each sample

Conos consumes per-sample objects that already have a PCA reduction. We build one with the pagoda2.1
constructor `Pagoda2$from()` and run only the reduction steps Conos needs — variance normalisation and PCA —
with `run(steps = ...)`. (Any `Pagoda2` or `Seurat` object carrying a reduction works.) `run()` is
thread/fork-safe, so set `n.cores` to your machine.

```{r}
#| message: false
#| warning: false
n.cores <- 4
preprocess <- function(cm) {
  rownames(cm) <- make.unique(rownames(cm))            # the panel has a few duplicate gene symbols
  p <- Pagoda2$from(cm, n.cores = n.cores, min.cells.per.gene = 0, min.transcripts.per.cell = 0,
                    log.scale = TRUE, verbose = FALSE)
  p$run(steps = c("variance", "pca"), dependencies = "error", verbose = FALSE)
  p
}
samples <- lapply(panel, preprocess)
```

## Build the Conos object and the joint graph

```{r}
con <- Conos$new(samples, n.cores = 1)
```

Before aligning, a quick sanity check of what the samples actually share. `planIntegration()` reports, per
modality, how many features are common across samples and how well they overlap — for this single-modality
(RNA) panel it simply confirms a large shared gene set:

```{r}
con$planIntegration()
```

`runGraph()` aligns every pair of samples and assembles the joint graph. The defaults are sensible:
`space = "PCA"` uses **reciprocal PCA** (fast and robust); `k` is the inter-sample neighbourhood size and
`k.self` the within-sample one.

```{r}
#| message: false
con$runGraph(k = 20, k.self = 5, space = "PCA", ncomps = 30, n.odgenes = 2000, verbose = FALSE)
```

## Cluster and embed

```{r}
#| message: false
con$runClustering(verbose = FALSE)                      # Leiden communities on the joint graph
con$runEmbedding(method = "largeVis", verbose = FALSE)  # a joint 2D embedding for visualisation
```

The joint embedding, coloured by the detected clusters:

```{r}
#| fig-width: 6
#| fig-height: 5
con$plotGraph(clustering = "leiden")
```

The same clusters laid over each sample separately. Because the clustering is *joint*, the same colours
appearing in every sample is the visual confirmation that populations are shared rather than sample-specific:

```{r}
#| fig-width: 8
#| fig-height: 6
con$plotPanel(clustering = "leiden", use.common.embedding = TRUE)
```

## Marker genes

`runMarkers()` finds genes differentially expressed between clusters; `plotMarkerDotPlot()` shows the top
few per cluster (dot size = fraction of cells expressing, colour = scaled mean expression):

```{r}
#| fig-width: 8
#| fig-height: 6
con$plotMarkerDotPlot(n.genes.per.group = 4)
```

## Transferring annotations

A common task is to carry annotations from one (or a few) labelled samples onto the rest. Here we *simulate*
that by hiding the cluster labels on a random 70% of cells and propagating from the remaining 30% over the
joint graph. `propagateLabels()` returns the transferred label plus a per-cell uncertainty.

```{r}
groups <- con$clusters$leiden$groups
set.seed(1)
reference <- groups[sample(length(groups), round(0.30 * length(groups)))]   # a small "annotated" subset
transfer  <- con$propagateLabels(labels = reference, verbose = FALSE)

# how often the transferred label matches the original on the held-out cells:
held.out <- setdiff(names(groups), names(reference))
mean(transfer$labels[held.out] == as.character(groups[held.out]))
```

`transfer$labels` is the per-cell label and `transfer$uncertainty` a 0–1 confidence you can threshold or
plot. On real data this recovers held-out labels with high accuracy.

## Recap

The whole standard workflow is four calls on the Conos object:

```r
con <- Conos$new(samples)
con$runGraph()        # align samples -> joint graph
con$runClustering()   # Leiden communities
con$runEmbedding()    # joint 2D embedding
```

plus `runMarkers()` / `plotMarkerDotPlot()` for markers and `propagateLabels()` for annotation transfer.
For controlling the alignment (different spaces, memory, multimodal facets) continue to the
[advanced tutorial](conos-advanced.ipynb).
