---
title: "Conos walkthrough: jointly analysing a panel of samples"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    self-contained: true
  ipynb: default
jupyter: ir44
---

**Conos** (Clustering On Network of Samples) analyses a *collection* of single-cell samples together.
Instead of merging all cells into one matrix, it aligns each pair of samples in a shared space, stitches
their neighbourhoods into one weighted graph, and runs clustering, embedding, and differential expression
on that joint graph. Cells of the same type connect across samples, while each sample keeps its own
structure — so you can ask "which populations are shared?", "how do their proportions differ between
conditions?", and "can I carry annotations from one sample to the rest?".

This walkthrough takes the standard workflow end to end on a small panel of bone-marrow (BM) and
cord-blood (CB) samples. For controlling the alignment (spaces, memory, multimodal data) see the
[advanced tutorial](conos-advanced.ipynb).

## Loading your data

A Conos panel is just a **named list of per-sample objects**, one per sample, each already processed with
[pagoda2](https://github.com/kharchenkolab/pagoda2) (or Seurat). In practice you build one object per
sample from its count data — for example:

```{r}
#| eval: false
library(pagoda2)

# from a 10x CellRanger output folder (contains matrix.mtx / barcodes / features):
s1 <- Pagoda2$from10x("sample1/outs/filtered_feature_bc_matrix")

# from a 10x HDF5 file:
s2 <- Pagoda2$from10xH5("sample2_filtered_feature_bc_matrix.h5")

# or directly from a counts matrix (genes x cells), e.g. one you loaded yourself:
s3 <- Pagoda2$from(my_counts_matrix)
```

**Cell names must be unique across the whole panel** so that no two cells (even in different samples) collide.

For this tutorial we use a ready-made example panel of four samples from the `conosPanel` package
(`install.packages("conosPanel", repos = "https://kharchenkolab.github.io/drat/")`). Each element is a
sparse counts matrix (genes × cells):

```{r}
#| message: false
#| warning: false
library(conos)
library(pagoda2)

panel <- conosPanel::panel
str(panel, 1)
```

To keep the tutorial quick we down-sample each sample to a few hundred cells; in a real analysis you would
use all of them.

```{r}
set.seed(1)
panel <- lapply(panel, function(m) m[, sample(ncol(m), 600)])
```

## Pre-processing each sample

Each sample is turned into a `Pagoda2` object carrying a PCA reduction. We wrap the two standard steps —
variance normalisation and PCA — in a small helper and apply it to every sample. (`Pagoda2$from()` accepts
a folder, an `.h5`, or a counts matrix as shown above.)

```{r}
#| message: false
#| warning: false
preprocess <- function(counts) {
  rownames(counts) <- make.unique(rownames(counts))         # de-duplicate any repeated gene symbols
  p <- Pagoda2$from(counts, n.cores = 1, min.cells.per.gene = 0,
                    min.transcripts.per.cell = 0, log.scale = TRUE, verbose = FALSE)
  p$run(steps = c("variance", "pca"), dependencies = "error", verbose = FALSE)
  p
}

samples <- lapply(panel, preprocess)
```

## Building the joint graph

Create the Conos object from the per-sample objects, then build the joint alignment graph. `runGraph()`
aligns every pair of samples on reciprocal PCA (`space = "PCA"`, the recommended default) and stitches the
neighbourhoods together; `k` / `k.self` set the inter- and within-sample neighbourhood sizes.

```{r}
con <- Conos$new(samples, n.cores = 1)
```

```{r}
#| message: false
con$runGraph(k = 15, k.self = 5, space = "PCA", ncomps = 30, n.odgenes = 2000, verbose = FALSE)
```

## Joint clustering

Detect clusters as communities on the joint graph (Leiden by default):

```{r}
#| message: false
con$runClustering(verbose = FALSE)
```

## Visualising the joint embedding

Compute a joint 2D embedding (UMAP) and colour it by the detected clusters:

```{r}
#| message: false
con$runEmbedding(method = "UMAP", min.visited.verts = 100, verbose = FALSE)
```

```{r}
#| fig-width: 6
#| fig-height: 5
con$plotGraph(clustering = "leiden")
```

## Are the samples well mixed?

The key question in any integration: do the samples *overlap* in the joint embedding (good — the same
cell types are aligned across samples), or do they sit apart (a sign of under-correction or genuinely
distinct populations)? Colour the same embedding by **sample** instead of cluster:

```{r}
#| fig-width: 6.5
#| fig-height: 5
con$plotGraph(color.by = "sample", mark.groups = FALSE, show.legend = TRUE, alpha = 0.1)
```

Cells from all four samples are interspersed within each region, so the populations are well aligned.

## Cluster composition across samples

`plotClusterBarplots()` shows what fraction of each cluster comes from each sample — useful for spotting
clusters that are sample- or condition-specific (here, differences between bone marrow and cord blood):

```{r}
#| fig-width: 7
#| fig-height: 4
plotClusterBarplots(con, clustering = "leiden", show.entropy = FALSE)
```

## A per-sample view

The same joint clustering laid over each sample separately. Because the clustering is *joint*, seeing the
same colours in every panel confirms the populations are shared rather than sample-specific:

```{r}
#| fig-width: 8
#| fig-height: 6
con$plotPanel(clustering = "leiden", use.common.embedding = TRUE)
```

## Marker genes

`runMarkers()` finds genes differentially expressed between clusters; `plotMarkerDotPlot()` shows the top
few per cluster (dot size = fraction of cells expressing, colour = scaled mean expression):

```{r}
#| fig-width: 8
#| fig-height: 6
con$plotMarkerDotPlot(n.genes.per.group = 3)
```

You can also paint a single gene onto the joint embedding:

```{r}
#| fig-width: 6
#| fig-height: 5
con$plotGraph(gene = "CD3E")
```

## Transferring annotations between samples

A common task is to carry cell-type labels from an annotated sample (or any labelled subset) to the rest.
`propagateLabels()` spreads the labels over the joint graph and returns, per cell, the transferred label
and an uncertainty score. Here we simulate it by labelling a random 40% of cells and propagating to the
rest:

```{r}
groups    <- con$clusters$leiden$groups
set.seed(1)
reference <- groups[sample(length(groups), round(0.40 * length(groups)))]
transfer  <- con$propagateLabels(labels = reference, verbose = FALSE)
```

```{r}
#| fig-width: 6
#| fig-height: 5
con$plotGraph(groups = transfer$labels)
```

`transfer$uncertainty` is a 0–1 score per cell (low = confident) that you can threshold or plot to flag
cells whose transferred label is ambiguous.

## Recap

The whole standard workflow is four calls on the Conos object:

```r
con <- Conos$new(samples)
con$runGraph()        # align samples -> joint graph
con$runClustering()   # Leiden communities
con$runEmbedding()    # joint 2D embedding
```

plus `plotGraph()` / `plotPanel()` / `plotClusterBarplots()` to visualise, `runMarkers()` /
`plotMarkerDotPlot()` for markers, and `propagateLabels()` for annotation transfer. To control the
alignment space, memory, and multimodal data, continue to the [advanced tutorial](conos-advanced.ipynb).
