---
title: "Conos advanced: alignment spaces, memory, modalities, label transfer"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    self-contained: true
  ipynb: default
jupyter: ir44
---

This tutorial assumes you have worked through the [walkthrough](conos-walkthrough.ipynb). It covers the
knobs that matter when integration is harder: **which space** to align in, **how strongly** to align,
**memory** on large panels, **what the samples actually share** (modalities), and how to **transfer
labels**. We reuse the same bone-marrow / cord-blood panel and take each topic one small step at a time.

## Contents

- [Setup](#setup)
- [Plan the integration first](#plan-the-integration-first)
- [Choosing the alignment space](#choosing-the-alignment-space)
- [Choosing a clustering resolution](#choosing-a-clustering-resolution)
- [Label transfer: diffusion vs solver](#label-transfer-diffusion-vs-solver)
- [Tuning the alignment strength](#tuning-the-alignment-strength)
- [Memory on large panels](#memory-on-large-panels)
- [Multimodal panels (facets)](#multimodal-panels-facets)
- [Recap](#recap)

## Setup

Same preparation as the walkthrough: load the panel, down-sample for speed, and turn each sample into a
`Pagoda2` object carrying a PCA reduction.

```{r}
#| message: false
#| warning: false
library(conos)
library(pagoda2)

set.seed(1)
panel <- lapply(conosPanel::panel, function(m) m[, sample(ncol(m), 600)])
```

```{r}
#| message: false
#| warning: false
samples <- lapply(panel, function(counts) {
  Pagoda2$from(counts, verbose = FALSE)$run(steps = c("variance", "pca"), verbose = FALSE)
})
con <- Conos$new(samples, n.cores = 1)
```

## Plan the integration first

`planIntegration()` is a read-only check you run **before** building the graph. It polls each sample's
molecular modalities (here just RNA; for multimodal samples it would also list ADT/ATAC) and, per modality,
reports how many features are shared across samples and how well they overlap — with a *usable / marginal /
not-integrable* verdict. This is where you catch problems like independently-called scATAC peaks that simply
cannot be reconciled.

```{r}
plan <- con$planIntegration()
plan
```

The resolved default modality is recorded on the object and is what `runGraph()` aligns on unless you tell
it otherwise.

## Choosing the alignment space

The `space` argument decides how each pair of samples is matched:

* **`"PCA"` (default)** — *reciprocal PCA*: each sample's PCA loadings define a shared gene basis that both
  samples project into. Fast and robust — the right default for most panels.
* **`"CCA"`** — canonical correlation. Useful for **heterogeneous** panels (very different compositions),
  but heavier and prone to **over-smoothing** (merging distinct populations) when little correction is
  actually needed.
* **`"CPCA"`** — common principal components; available, but usually much slower without a clear gain over
  reciprocal PCA, so it is not the default.

Rather than take that on faith, let's see it. We build the graph with each space and look at the joint
embedding **coloured by sample** — the more the samples interleave, the more aggressively that space mixed
them. First the default, reciprocal PCA:

```{r}
#| message: false
con$runGraph(space = "PCA", k = 20, k.self = 5, ncomps = 30, n.odgenes = 2000, verbose = FALSE)
con$runEmbedding(method = "UMAP", embedding.name = "umap_pca", min.visited.verts = 100, verbose = FALSE)
```

```{r}
#| fig-width: 6.5
#| fig-height: 5
con$plotGraph(embedding = "umap_pca", color.by = "sample", mark.groups = FALSE, show.legend = TRUE, alpha = 0.1)
```

Now CCA, on the same panel (a separately named embedding so we can compare the two):

```{r}
#| message: false
con$runGraph(space = "CCA", k = 20, k.self = 5, ncomps = 30, n.odgenes = 2000, verbose = FALSE)
con$runEmbedding(method = "UMAP", embedding.name = "umap_cca", min.visited.verts = 100, verbose = FALSE)
```

```{r}
#| fig-width: 6.5
#| fig-height: 5
con$plotGraph(embedding = "umap_cca", color.by = "sample", mark.groups = FALSE, show.legend = TRUE, alpha = 0.1)
```

CCA pulls the samples together more tightly. On a well-behaved panel like this one that is unnecessary —
reciprocal PCA already interleaves the samples — and over-mixing risks merging populations that should stay
separate. So we **keep reciprocal PCA** for the rest of the tutorial; rebuild the graph and embedding we'll
use from here on:

```{r}
#| message: false
con$runGraph(space = "PCA", k = 20, k.self = 5, ncomps = 30, n.odgenes = 2000, verbose = FALSE)
con$runEmbedding(method = "UMAP", min.visited.verts = 100, verbose = FALSE)
```

## Choosing a clustering resolution

`scanResolution()` runs the community detection over a range of resolutions and reports the number of
clusters (and modularity), so you can pick a value deliberately rather than guess:

```{r}
#| message: false
#| fig-width: 6
#| fig-height: 4
res <- scanResolution(con, resolutions = c(0.5, 1, 1.5, 2), plot = TRUE, verbose = FALSE)
res
```

Higher resolution = more, finer clusters. Pick the coarsest resolution that still separates the populations
you care about, then commit to it:

```{r}
#| message: false
con$runClustering(resolution = 1.5, name = "leiden_r15", verbose = FALSE)
```

## Label transfer: diffusion vs solver

`propagateLabels()` spreads a partial set of labels over the joint graph. It has two engines that give
near-identical results — the difference is how they scale. Let's walk through it.

First set up the problem: take the clustering as ground truth, reveal a random **30%** of cells as the
"reference", and hide the rest.

```{r}
groups    <- con$clusters$leiden_r15$groups
set.seed(1)
reference <- groups[sample(length(groups), round(0.30 * length(groups)))]
held.out  <- setdiff(names(groups), names(reference))
```

The reference is sparse — only a scattering of labelled seeds across the embedding:

```{r}
#| fig-width: 6
#| fig-height: 5
con$plotGraph(groups = reference, mark.groups = FALSE, show.legend = TRUE)
```

**`"diffusion"` (default)** — iterative graph diffusion; scales to very large panels.

```{r}
#| message: false
td <- con$propagateLabels(labels = reference, method = "diffusion", verbose = FALSE)
```

```{r}
#| fig-width: 6
#| fig-height: 5
con$plotGraph(groups = td$labels)
```

The labels have filled in across the whole embedding from those few seeds.

**`"solver"`** — a Zhu–Ghahramani–Lafferty harmonic solve; fast and slightly more accurate at typical
sizes, but its linear solve does not scale much past ~20k cells.

```{r}
#| message: false
ts <- con$propagateLabels(labels = reference, method = "solver", solver = "Matrix")
```

Both engines recover the held-out labels well above chance, and they largely agree with each other:

```{r}
c(diffusion_accuracy = round(mean(td$labels[held.out] == as.character(groups[held.out])), 3),
  solver_accuracy    = round(mean(ts$labels[held.out] == as.character(groups[held.out])), 3),
  engine_agreement   = round(mean(td$labels[held.out] == ts$labels[held.out]), 3))
```

Either way, `*$uncertainty` flags the cells whose transferred label is least trustworthy — typically those
at the boundaries between populations. Painting it onto the embedding tells you *where* to trust the
transfer:

```{r}
#| fig-width: 6
#| fig-height: 5
con$plotGraph(colors = td$uncertainty)
```

Use `diffusion` (the default) for large atlases, and `solver` when the panel is modest and you want the
extra speed/accuracy.

## Tuning the alignment strength

Within a space, **how forcefully** samples are pulled together is set by `alignment.strength` (0 = none,
higher = stronger) together with the neighbourhood sizes `k` / `k.self`. Raise it for stronger batch
correction across heterogeneous samples; lower it to preserve subtle within-sample structure. It is the
first dial to reach for when the sample-mixing plot looks under- or over-corrected — leave it at the default
otherwise.

```{r}
#| message: false
con$runGraph(space = "PCA", alignment.strength = 0.3, k = 20, k.self = 5,
             ncomps = 30, n.odgenes = 2000, verbose = FALSE)
```

(After changing the graph you would re-run `runClustering()` / `runEmbedding()` to see the effect.)

## Memory on large panels

The per-pair alignment rotations are O(samples²) and dominate memory on big panels. `pairs.storage`
controls where they live:

* **`"keep"` (default)** — kept in RAM for reuse.
* **`"drop"`** — discarded after the graph is built (a later `runGraph()` recomputes them).
* **`"disk"`** — offloaded to a scratch file: freed from RAM, but restored on the next `runGraph()` so
  re-runs stay cheap.

The disk path costs almost nothing in time, because the rotations are written once after the graph is built.
We can time it:

```{r}
#| message: false
t_keep <- system.time(con$runGraph(space = "PCA", k = 20, k.self = 5, ncomps = 30, n.odgenes = 2000,
                                   pairs.storage = "keep", verbose = FALSE))["elapsed"]
t_disk <- system.time(con$runGraph(space = "PCA", k = 20, k.self = 5, ncomps = 30, n.odgenes = 2000,
                                   pairs.storage = "disk", verbose = FALSE))["elapsed"]
c(keep_seconds = round(t_keep, 2), disk_seconds = round(t_disk, 2))
```

On a four-sample panel the RAM saved is small, but it grows as the *square* of the number of samples — so
`"disk"` is the right choice for large atlases, at essentially no time cost. (On lstar-backed pagoda2
samples the build path also streams expression in gene blocks, keeping the per-sample footprint flat.)

## Multimodal panels (facets)

When samples carry more than one molecular modality (RNA + ADT in CITE-seq, or RNA + ATAC in multiome,
produced with pagoda2 *facets*), `planIntegration()` polls and assesses **each** modality separately, so you
can see which ones are actually integrable across the panel before committing. Today Conos aligns on a
single shared modality (the common default, usually RNA) — the recommended, robust path — and
modality-aware accessors (`getModalities()`, `getDefaultModality()`) expose the rest. Joint multimodal
*fusion* (weighting modalities per cell, à la WNN) is the next step on the roadmap.

## Recap

`planIntegration()` to see what's realistic → `runGraph(space=, alignment.strength=, pairs.storage=)` to
control the alignment and memory → `scanResolution()` + `runClustering(resolution=)` to tune granularity →
`propagateLabels(method=)` to transfer annotations. Start from the
[walkthrough](conos-walkthrough.ipynb) defaults and reach for these knobs only when a particular integration
needs them.
