---
title: "Conos data sources: Seurat objects, anndata / Seurat / loom files, and lstar stores"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    self-contained: true
  ipynb: default
jupyter: ir44
---

A Conos panel is just a **named list of per-sample objects**, and those samples do not all have to be
pagoda2 objects built in this session. Conos works with:

* **pagoda2** objects (the [walkthrough](conos-walkthrough.ipynb) default),
* **Seurat** objects in memory (v3, v4, or v5),
* samples **read from files** written by other pipelines — **anndata / scanpy** (`.h5ad`), **Seurat**
  (`.h5seurat`), the **loom** format, or an **lstar** (zarr) store,
* and even a **mix** of object types in one panel.

pagoda2's `from*()` constructors read each file format directly (via HDF5 / zarr — no Python required) into
an object Conos can align. This tutorial shows each route on the bone-marrow / cord-blood panel.

## At a glance

```r
library(conos)
library(pagoda2)

# Seurat objects in memory work directly:
con <- Conos$new(list(a = seurat_a, b = seurat_b))

# ... or read each sample from a file (each returns a Pagoda2 object; run() adds the PCA Conos aligns on):
s1 <- Pagoda2$fromAnnData("sample1.h5ad")$run(steps = c("variance", "pca"))     # anndata / scanpy
s2 <- Pagoda2$fromH5Seurat("sample2.h5seurat")$run(steps = c("variance", "pca")) # Seurat (SeuratDisk)
s3 <- Pagoda2$fromLoom("sample3.loom")$run(steps = c("variance", "pca"))         # loom
s4 <- Pagoda2$fromLstar("sample4.zarr")$run(steps = c("variance", "pca"))        # lstar (zarr) store
con <- Conos$new(list(s1, s2, s3, s4))

# from here everything is identical to the walkthrough:
con$runGraph(); con$runClustering(); con$runEmbedding()

# the WHOLE collection (panel + joint graph/clusters/embedding) also round-trips and exports:
lstar::write_conos(con) |> lstar::lstar_write("panel.lstar.zarr")  # save the whole Conos (lstar zarr)
con <- lstar::read_conos("panel.lstar.zarr")                       # ... and reopen it

so  <- lstar::write_seurat(lstar::write_conos(con))   # -> Seurat v5 (split assay + joint graph + embedding)
con <- lstar::read_conos(lstar::read_seurat(so))      # ... and read a collection back from it
```

## Contents

- [Example data](#example-data)
- [Seurat objects in memory](#seurat-objects-in-memory)
- [Reading samples from files](#reading-samples-from-files)
- [Exporting a whole collection](#exporting-a-whole-collection)
- [Mixing object types in one panel](#mixing-object-types-in-one-panel)

## Example data

We reuse the `conosPanel` matrices and down-sample them for speed. These raw count matrices stand in for
"data you have on disk"; below we will turn them into Seurat objects and into files written in several
formats, then read them back.

```{r}
#| message: false
#| warning: false
library(conos)
library(pagoda2)

set.seed(1)
raw <- lapply(conosPanel::panel, function(m) {
  m <- m[, sample(ncol(m), 700)]
  rownames(m) <- make.unique(rownames(m)); colnames(m) <- make.unique(colnames(m))
  m
})
str(raw, 1)
```

## Seurat objects in memory

If your samples are already `Seurat` objects, hand them straight to `Conos$new()` — no conversion needed.
Build each one the usual Seurat way (normalise, find variable features, scale, PCA); Conos aligns on the
PCA. We do two samples here:

```{r}
#| message: false
#| warning: false
library(Seurat)

seurat.samples <- lapply(raw[1:2], function(cm) {
  s <- CreateSeuratObject(counts = cm)
  s <- NormalizeData(s, verbose = FALSE)
  s <- FindVariableFeatures(s, nfeatures = 2000, verbose = FALSE)
  s <- ScaleData(s, verbose = FALSE)
  RunPCA(s, npcs = 30, verbose = FALSE)
})
```

```{r}
#| message: false
#| warning: false
con.seurat <- Conos$new(seurat.samples, n.cores = 1)
con.seurat$runGraph(k = 15, k.self = 5, space = "PCA", ncomps = 30, n.odgenes = 2000, verbose = FALSE)
con.seurat$runClustering(verbose = FALSE)
con.seurat$runEmbedding(method = "UMAP", min.visited.verts = 100, verbose = FALSE)
```

The joint embedding, coloured by sample, shows the two Seurat samples are well mixed — exactly as for a
pagoda2 panel:

```{r}
#| fig-width: 6.5
#| fig-height: 5
con.seurat$plotGraph(color.by = "sample", mark.groups = FALSE, alpha = 0.1, show.legend = TRUE)
```

Clustering, markers, embedding and label transfer all work on Seurat panels too:

```{r}
#| fig-width: 9
#| fig-height: 5
con.seurat$plotMarkerDotPlot(n.genes.per.group = 3)
```

## Reading samples from files

More often, each sample lives in a file produced by another pipeline. pagoda2's `from*()` constructors read
each format directly (HDF5 / zarr — no Python). To make this runnable we first **write** three samples (in
practice they already exist); we use the anndata `.h5ad` and lstar (zarr) formats, both written by pagoda2,
which preserve gene names exactly:

```{r}
#| message: false
#| warning: false
dir <- file.path(tempdir(), "conos-sources"); dir.create(dir, showWarnings = FALSE)
paths <- c(bm1 = file.path(dir, "bm1.h5ad"),
           bm2 = file.path(dir, "bm2.lstar.zarr"),
           cb1 = file.path(dir, "cb1.h5ad"))
Pagoda2$from(raw[[1]], verbose = FALSE)$export(paths[["bm1"]], format = "h5ad",  overwrite = TRUE)
Pagoda2$from(raw[[2]], verbose = FALSE)$export(paths[["bm2"]], format = "lstar", overwrite = TRUE)
Pagoda2$from(raw[[3]], verbose = FALSE)$export(paths[["cb1"]], format = "h5ad",  overwrite = TRUE)
```

A Seurat `.h5seurat` and a `.loom` file are read exactly the same way — point the matching constructor at
the file:

```{r}
#| eval: false
s_h5seurat <- Pagoda2$fromH5Seurat("sample.h5seurat")$run(steps = c("variance", "pca"))
s_loom     <- Pagoda2$fromLoom("sample.loom")$run(steps = c("variance", "pca"))
```

(`fromH5Seurat()` reads any valid `.h5seurat`; if a file lacks stored gene names — e.g. one written from a
Seurat v5 object by a `SeuratObject < 5` build of SeuratDisk, which can't serialize `Assay5` feature names
— pagoda2 **warns** and falls back to positional names rather than silently mislabelling genes.)

Read each file with the matching constructor and add the variance + PCA reduction Conos needs — the result
is an ordinary Conos panel:

```{r}
#| message: false
#| warning: false
samples <- list(
  bm1 = Pagoda2$fromAnnData(paths[["bm1"]])$run(steps = c("variance", "pca"), verbose = FALSE),
  bm2 = Pagoda2$fromLstar(paths[["bm2"]])$run(steps = c("variance", "pca"), verbose = FALSE),
  cb1 = Pagoda2$fromAnnData(paths[["cb1"]])$run(steps = c("variance", "pca"), verbose = FALSE)
)
sapply(samples, function(s) length(getCellNames(s)))
```

These read-from-disk samples integrate exactly like any other:

```{r}
#| message: false
#| warning: false
con.files <- Conos$new(samples, n.cores = 1)
con.files$runGraph(k = 15, k.self = 5, space = "PCA", ncomps = 30, n.odgenes = 2000, verbose = FALSE)
con.files$runClustering(verbose = FALSE)
con.files$runEmbedding(method = "UMAP", min.visited.verts = 100, verbose = FALSE)
```

```{r}
#| fig-width: 6.5
#| fig-height: 5
con.files$plotGraph(color.by = "sample", mark.groups = FALSE, alpha = 0.1, show.legend = TRUE)
```

## Exporting a whole collection

A `Conos` object is itself a **collection**: every sample's raw counts *plus* the joint analysis layer —
the graph, the embedding and the clustering — over the union of cells. Conos integrates in *graph* space,
so there is **no batch-corrected expression matrix** (computing one is expensive and lossy). lstar writes
the whole collection to the native multi-sample formats *without fabricating* such a matrix: each sample's
raw counts are preserved, and the joint graph / embedding / clusters travel alongside.

### An lstar (zarr) store

The native round-trip. `write_conos()` turns the `Conos` into an lstar collection, `lstar_write()` saves it,
and `read_conos()` reopens it as a live `Conos` — the per-sample objects and the joint graph / clustering /
embedding are all restored, so you can carry on plotting, finding markers and transferring labels without
recomputing the integration:

```{r}
#| message: false
library(lstar)
collection.path <- file.path(dir, "panel.lstar.zarr")
lstar_write(write_conos(con.files), collection.path)

con.reopened <- read_conos(collection.path)
c(samples  = length(con.reopened$samples),
  cells    = length(con.reopened$clusters$leiden$groups),
  clusters = nlevels(con.reopened$clusters$leiden$groups))
```

```{r}
#| fig-width: 6
#| fig-height: 5
con.reopened$plotGraph(clustering = "leiden")
```

### A Seurat v5 object

The same collection converts to a **Seurat v5** object: a *split* assay with one raw-count layer per
sample, the joint graph as a Seurat `Graph`, the embedding as a `DimReduc`, and the sample / cluster labels
in `meta.data`. `write_seurat()` takes the lstar collection, so chain it after `write_conos()`:

```{r}
#| message: false
#| warning: false
so <- write_seurat(write_conos(con.files))
c(class     = class(so)[1],
  layers    = length(SeuratObject::Layers(so, assay = "RNA")),   # one raw-count layer per sample
  graph     = paste(SeuratObject::Graphs(so), collapse = ","),    # joint graph, carried natively
  reduction = paste(SeuratObject::Reductions(so), collapse = ","))
saveRDS(so, file.path(dir, "panel_seurat_v5.rds"))
```

You can also read a collection **back from such a Seurat v5 `.rds`**: `read_seurat()` recovers the lstar
collection and `read_conos()` rebuilds the live `Conos` — joint graph and all — so a Seurat-v5 integration
object opens directly as a `Conos`:

```{r}
#| message: false
con.from.seurat <- read_conos(read_seurat(readRDS(file.path(dir, "panel_seurat_v5.rds"))))
c(samples = length(con.from.seurat$samples), has.graph = !is.null(con.from.seurat$graph))
```

### An AnnData file (for scanpy)

For scanpy users the stored collection flattens to a single AnnData: `X` holds the **raw joint counts**
(again, no corrected matrix), the joint graph lands in `obsp` (aliased to `connectivities`, with a matching
`uns['neighbors']`) so scanpy's graph tools run on the conos graph directly, the embedding in `obsm`, and
sample / cluster in `obs`. Convert the store from the shell with the lstar CLI:

```bash
lstar convert panel.lstar.zarr panel.h5ad
```

(The same `lstar convert` command also moves *individual* objects between formats — e.g.
`lstar convert sample.rds sample.h5ad` — so lstar doubles as a general format converter between Seurat,
SingleCellExperiment, anndata and zarr.)

### Running conos from a Python (scanpy) workflow

Conos itself is R-only, but the two directions above add up to a clean recommendation for **scanpy users**:
keep your data as `.h5ad` and bridge through one short R step. **First**, the conos integration — the only R
in the loop, saved as `conos_step.R` and run non-interactively:

```{r}
#| eval: false
# conos_step.R
library(conos); library(pagoda2); library(lstar)
samples <- lapply(c("s1.h5ad", "s2.h5ad", "s3.h5ad"),
                  function(f) Pagoda2$fromAnnData(f)$run(steps = c("variance", "pca")))
con <- Conos$new(samples)
con$runGraph(); con$runClustering(); con$runEmbedding()
lstar_write(write_conos(con), "panel.lstar.zarr")
```

**Then**, run it and convert the collection to a scanpy-ready `.h5ad` — both from the shell:

```bash
Rscript conos_step.R                               # the conos integration (no interactive R)
lstar convert panel.lstar.zarr integrated.h5ad     # -> a scanpy-ready .h5ad
```

**Finally**, pick up in scanpy. The conos joint graph is already in `adata.obsp['connectivities']`, so
clustering and embedding run directly on it — no re-integration:

```python
import scanpy as sc
adata = sc.read_h5ad("integrated.h5ad")
sc.tl.leiden(adata)   # runs on the conos joint graph in obsp['connectivities']
sc.tl.umap(adata)     # (conos's own embedding is already in adata.obsm)
```

There is no pure-Python conos build — the R step *is* the integration. If you must stay entirely in Python,
a Python-native method (scVI, Harmony, BBKNN) is the alternative, but it won't reproduce conos's joint-graph
alignment.

## Mixing object types in one panel

A panel may combine sources — say one sample is already a `Pagoda2` object and another is a `Seurat`
object. Conos aligns them together. Markers are computed **consistently** across the panel: pagoda2 samples
use pagoda2's own (disk-backed) differential-expression path, and Seurat (or other) samples are scored with
the *same* Wilcoxon test via `sccore`, so the result does not depend on how each sample was pre-processed.

```{r}
#| message: false
#| warning: false
mixed <- list(
  bm_pagoda2 = Pagoda2$from(raw[[2]], verbose = FALSE)$run(steps = c("variance", "pca"), verbose = FALSE),
  cb_seurat  = RunPCA(ScaleData(FindVariableFeatures(NormalizeData(CreateSeuratObject(counts = raw[[4]]),
                       verbose = FALSE), nfeatures = 2000, verbose = FALSE), verbose = FALSE),
                      npcs = 30, verbose = FALSE)
)
sapply(mixed, function(s) class(s)[1])
```

```{r}
#| message: false
#| warning: false
con.mixed <- Conos$new(mixed, n.cores = 1)
con.mixed$runGraph(k = 15, k.self = 5, space = "PCA", ncomps = 30, n.odgenes = 2000, verbose = FALSE)
con.mixed$runClustering(verbose = FALSE)
con.mixed$runEmbedding(method = "UMAP", min.visited.verts = 100, verbose = FALSE)
```

The pagoda2 and Seurat samples interleave in the joint embedding:

```{r}
#| fig-width: 6.5
#| fig-height: 5
con.mixed$plotGraph(color.by = "sample", mark.groups = FALSE, alpha = 0.1, show.legend = TRUE)
```

And markers — computed identically for both sample types — give a clean per-cluster picture:

```{r}
#| fig-width: 9
#| fig-height: 5
con.mixed$plotMarkerDotPlot(n.genes.per.group = 3)
```

Whatever the source — pagoda2 or Seurat, in memory or on disk — once the samples are in a list, the rest of
the analysis is the [walkthrough](conos-walkthrough.ipynb), and the alignment knobs are in the
[advanced tutorial](conos-advanced.ipynb).
