---
title: "Conos data sources: Seurat objects, anndata / Seurat / loom files, and lstar stores"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    self-contained: true
  ipynb: default
jupyter: ir44
---

A Conos panel is just a **named list of per-sample objects**, and those samples do not all have to be
pagoda2 objects built in this session. Conos works with:

* **pagoda2** objects (the [walkthrough](conos-walkthrough.ipynb) default),
* **Seurat** objects in memory (v3, v4, or v5),
* samples **read from files** written by other pipelines — **anndata / scanpy** (`.h5ad`), **Seurat**
  (`.h5seurat`), the **loom** format, or an **lstar** (zarr) store,
* and even a **mix** of object types in one panel.

pagoda2's `from*()` constructors read each file format directly (via HDF5 / zarr — no Python required) into
an object Conos can align. This tutorial shows each route on the bone-marrow / cord-blood panel.

## At a glance

```r
library(conos)
library(pagoda2)

# Seurat objects in memory work directly:
con <- Conos$new(list(a = seurat_a, b = seurat_b))

# ... or read each sample from a file (each returns a Pagoda2 object; run() adds the PCA Conos aligns on):
s1 <- Pagoda2$fromAnnData("sample1.h5ad")$run(steps = c("variance", "pca"))     # anndata / scanpy
s2 <- Pagoda2$fromH5Seurat("sample2.h5seurat")$run(steps = c("variance", "pca")) # Seurat (SeuratDisk)
s3 <- Pagoda2$fromLoom("sample3.loom")$run(steps = c("variance", "pca"))         # loom
s4 <- Pagoda2$fromLstar("sample4.zarr")$run(steps = c("variance", "pca"))        # lstar (zarr) store
con <- Conos$new(list(s1, s2, s3, s4))

# from here everything is identical to the walkthrough:
con$runGraph(); con$runClustering(); con$runEmbedding()

# an lstar store can also hold a WHOLE collection (panel + joint graph/clusters/embedding):
lstar::write_conos(con) |> lstar::lstar_write("panel.lstar.zarr")  # save the whole Conos
con <- lstar::read_conos("panel.lstar.zarr")                       # ... and reopen it
```

## Contents

- [Example data](#example-data)
- [Seurat objects in memory](#seurat-objects-in-memory)
- [Reading samples from files](#reading-samples-from-files)
- [An entire collection in one store](#an-entire-collection-in-one-store)
- [Mixing object types in one panel](#mixing-object-types-in-one-panel)

## Example data

We reuse the `conosPanel` matrices and down-sample them for speed. These raw count matrices stand in for
"data you have on disk"; below we will turn them into Seurat objects and into files written in several
formats, then read them back.

```{r}
#| message: false
#| warning: false
library(conos)
library(pagoda2)

set.seed(1)
raw <- lapply(conosPanel::panel, function(m) {
  m <- m[, sample(ncol(m), 700)]
  rownames(m) <- make.unique(rownames(m)); colnames(m) <- make.unique(colnames(m))
  m
})
str(raw, 1)
```

## Seurat objects in memory

If your samples are already `Seurat` objects, hand them straight to `Conos$new()` — no conversion needed.
Build each one the usual Seurat way (normalise, find variable features, scale, PCA); Conos aligns on the
PCA. We do two samples here:

```{r}
#| message: false
#| warning: false
library(Seurat)

seurat.samples <- lapply(raw[1:2], function(cm) {
  s <- CreateSeuratObject(counts = cm)
  s <- NormalizeData(s, verbose = FALSE)
  s <- FindVariableFeatures(s, nfeatures = 2000, verbose = FALSE)
  s <- ScaleData(s, verbose = FALSE)
  RunPCA(s, npcs = 30, verbose = FALSE)
})
```

```{r}
#| message: false
con.seurat <- Conos$new(seurat.samples, n.cores = 1)
con.seurat$runGraph(k = 15, k.self = 5, space = "PCA", ncomps = 30, n.odgenes = 2000, verbose = FALSE)
con.seurat$runClustering(verbose = FALSE)
con.seurat$runEmbedding(method = "UMAP", min.visited.verts = 100, verbose = FALSE)
```

The joint embedding, coloured by sample, shows the two Seurat samples are well mixed — exactly as for a
pagoda2 panel:

```{r}
#| fig-width: 6.5
#| fig-height: 5
con.seurat$plotGraph(color.by = "sample", mark.groups = FALSE, alpha = 0.1, show.legend = TRUE)
```

Clustering, markers, embedding and label transfer all work on Seurat panels too:

```{r}
#| fig-width: 9
#| fig-height: 5
con.seurat$plotMarkerDotPlot(n.genes.per.group = 3)
```

## Reading samples from files

More often, each sample lives in a file produced by another pipeline. To make this runnable we first
**write** three samples to three different formats (in practice they already exist): an anndata `.h5ad`, an
lstar (zarr) store, and a Seurat `.h5seurat`.

```{r}
#| message: false
#| warning: false
library(SeuratDisk)
dir <- file.path(tempdir(), "conos-sources"); dir.create(dir, showWarnings = FALSE)
paths <- c(bm1 = file.path(dir, "bm1.h5ad"),
           bm2 = file.path(dir, "bm2.lstar.zarr"),
           cb1 = file.path(dir, "cb1.h5seurat"))

# anndata + lstar are written by pagoda2:
Pagoda2$from(raw[[1]], verbose = FALSE)$export(paths[["bm1"]], format = "h5ad",  overwrite = TRUE)
Pagoda2$from(raw[[2]], verbose = FALSE)$export(paths[["bm2"]], format = "lstar", overwrite = TRUE)

# a Seurat .h5seurat, written by SeuratDisk. NB: writing .h5seurat under Seurat v5 needs a
# SeuratObject>=5-compatible SeuratDisk, and a v3 `Assay` (Assay5 has no h5seurat writer yet):
old.opt <- options(Seurat.object.assay.version = "v3")
cb1.seurat <- NormalizeData(CreateSeuratObject(counts = raw[[3]]), verbose = FALSE)
SaveH5Seurat(cb1.seurat, filename = paths[["cb1"]], overwrite = TRUE, verbose = FALSE)
options(old.opt)
```

Read each file with the matching `Pagoda2$from*()` constructor and add the variance + PCA reduction Conos
needs — the result is an ordinary Conos panel. (A `.loom` file reads the same way with
`Pagoda2$fromLoom()`.)

```{r}
#| message: false
#| warning: false
samples <- list(
  bm1 = Pagoda2$fromAnnData(paths[["bm1"]])$run(steps = c("variance", "pca"), verbose = FALSE),
  bm2 = Pagoda2$fromLstar(paths[["bm2"]])$run(steps = c("variance", "pca"), verbose = FALSE),
  cb1 = Pagoda2$fromH5Seurat(paths[["cb1"]])$run(steps = c("variance", "pca"), verbose = FALSE)
)
sapply(samples, function(s) length(getCellNames(s)))
```

These read-from-disk samples integrate exactly like any other:

```{r}
#| message: false
con.files <- Conos$new(samples, n.cores = 1)
con.files$runGraph(k = 15, k.self = 5, space = "PCA", ncomps = 30, n.odgenes = 2000, verbose = FALSE)
con.files$runClustering(verbose = FALSE)
con.files$runEmbedding(method = "UMAP", min.visited.verts = 100, verbose = FALSE)
```

```{r}
#| fig-width: 6.5
#| fig-height: 5
con.files$plotGraph(color.by = "sample", mark.groups = FALSE, alpha = 0.1, show.legend = TRUE)
```

## An entire collection in one store

An lstar (zarr) store need not hold a single sample — it can hold a **whole collection**: every sample
*plus* the joint graph, clustering and embedding. That is conos's level, so conos writes the entire object
and reads it straight back. Save the panel we just built into one store:

```{r}
#| message: false
library(lstar)
collection.path <- file.path(dir, "panel.lstar.zarr")
lstar_write(write_conos(con.files), collection.path)
```

`read_conos()` reopens it as a live `Conos` — the per-sample objects and the joint graph / clustering /
embedding are all restored, so you can carry on plotting, finding markers and transferring labels without
recomputing the integration:

```{r}
#| message: false
con.reopened <- read_conos(collection.path)
c(samples  = length(con.reopened$samples),
  cells    = length(con.reopened$clusters$leiden$groups),
  clusters = nlevels(con.reopened$clusters$leiden$groups))
```

```{r}
#| fig-width: 6
#| fig-height: 5
con.reopened$plotGraph(clustering = "leiden")
```

## Mixing object types in one panel

A panel may combine sources — say one sample is already a `Pagoda2` object and another is a `Seurat`
object. Conos aligns them together. Markers are computed **consistently** across the panel: pagoda2 samples
use pagoda2's own (disk-backed) differential-expression path, and Seurat (or other) samples are scored with
the *same* Wilcoxon test via `sccore`, so the result does not depend on how each sample was pre-processed.

```{r}
#| message: false
#| warning: false
mixed <- list(
  bm_pagoda2 = Pagoda2$from(raw[[2]], verbose = FALSE)$run(steps = c("variance", "pca"), verbose = FALSE),
  cb_seurat  = RunPCA(ScaleData(FindVariableFeatures(NormalizeData(CreateSeuratObject(counts = raw[[4]]),
                       verbose = FALSE), nfeatures = 2000, verbose = FALSE), verbose = FALSE),
                      npcs = 30, verbose = FALSE)
)
sapply(mixed, function(s) class(s)[1])
```

```{r}
#| message: false
con.mixed <- Conos$new(mixed, n.cores = 1)
con.mixed$runGraph(k = 15, k.self = 5, space = "PCA", ncomps = 30, n.odgenes = 2000, verbose = FALSE)
con.mixed$runClustering(verbose = FALSE)
con.mixed$runEmbedding(method = "UMAP", min.visited.verts = 100, verbose = FALSE)
```

The pagoda2 and Seurat samples interleave in the joint embedding:

```{r}
#| fig-width: 6.5
#| fig-height: 5
con.mixed$plotGraph(color.by = "sample", mark.groups = FALSE, alpha = 0.1, show.legend = TRUE)
```

And markers — computed identically for both sample types — give a clean per-cluster picture:

```{r}
#| fig-width: 9
#| fig-height: 5
con.mixed$plotMarkerDotPlot(n.genes.per.group = 3)
```

Whatever the source — pagoda2 or Seurat, in memory or on disk — once the samples are in a list, the rest of
the analysis is the [walkthrough](conos-walkthrough.ipynb), and the alignment knobs are in the
[advanced tutorial](conos-advanced.ipynb).
