---
title: "Conos: adjusting alignment strength"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    self-contained: true
  ipynb: default
jupyter: ir44
---

By default Conos aligns samples gently — it corrects obvious batch differences while preserving each
sample's structure. Sometimes that isn't enough: samples from different tissues or conditions still form
**sample-specific sub-clusters** within a shared cell type. When that residual separation is batch rather
than biology, you can pull the samples together more forcefully. This tutorial covers the two knobs for
that — the `alignment.strength` dial and **"supervised" alignment** that down-weights edges within a chosen
factor — on the bone-marrow (BM) / cord-blood (CB) panel.

A word of caution up front: stronger alignment trades structure for mixing. Push it too far and genuinely
distinct populations get merged. Always sanity-check that the populations you care about stay separated.

## At a glance

```r
# pull samples together more forcefully (0 = none ... 1 = maximum):
con$runGraph(alignment.strength = 0.3)

# or align only *across* a chosen factor (e.g. tissue), down-weighting within-factor edges:
con$runGraph(balancing.factor.per.cell = tissue, same.factor.downweight = 0.1, alignment.strength = 0.3)
```

## Contents

- [Setup](#setup)
- [A baseline alignment](#a-baseline-alignment)
- [Turning up the alignment strength](#turning-up-the-alignment-strength)
- [Supervised alignment](#supervised-alignment)
- [Choosing a value](#choosing-a-value)

## Setup

Load the panel, down-sample for speed, and pre-process each sample into a `Pagoda2` object with a PCA
reduction (as in the [walkthrough](conos-walkthrough.ipynb)).

```{r}
#| message: false
#| warning: false
library(conos)
library(pagoda2)

set.seed(1)
panel <- lapply(conosPanel::panel, function(m) m[, sample(ncol(m), 600)])
samples <- lapply(panel, function(cm) {
  Pagoda2$from(cm, verbose = FALSE)$run(steps = c("variance", "pca"), verbose = FALSE)
})
con <- Conos$new(samples, n.cores = 1)
```

## A baseline alignment

Build the joint graph with the defaults and embed it. Colouring the embedding by **sample** is how we judge
alignment: well-aligned samples interleave; residual batch shows up as sample-specific regions.

```{r}
#| message: false
con$runGraph(k = 20, k.self = 5, space = "PCA", ncomps = 30, verbose = FALSE)
con$runEmbedding(method = "UMAP", embedding.name = "default", min.visited.verts = 100, verbose = FALSE)
```

```{r}
#| fig-width: 6.5
#| fig-height: 5
con$plotGraph(embedding = "default", color.by = "sample", mark.groups = FALSE, alpha = 0.1, show.legend = TRUE)
```

The four samples mostly overlap, but some regions are dominated by one tissue — the kind of residual
separation `alignment.strength` can reduce.

## Turning up the alignment strength

`alignment.strength` runs from 0 (the gentle default behaviour) to 1 (maximum). Higher values pull mutual
neighbours across samples together more aggressively. Rebuild the graph at a moderate value and re-embed
(into a separately named embedding so we can compare):

```{r}
#| message: false
con$runGraph(k = 20, k.self = 5, space = "PCA", ncomps = 30, alignment.strength = 0.3, verbose = FALSE)
con$runEmbedding(method = "UMAP", embedding.name = "strength_0.3", min.visited.verts = 100, verbose = FALSE)
```

```{r}
#| fig-width: 6.5
#| fig-height: 5
con$plotGraph(embedding = "strength_0.3", color.by = "sample", mark.groups = FALSE, alpha = 0.1, show.legend = TRUE)
```

The samples are noticeably more interleaved. Pushing further (e.g. `alignment.strength = 0.7`) mixes them
even more tightly — but watch for distinct populations starting to collapse together:

```{r}
#| message: false
con$runGraph(k = 20, k.self = 5, space = "PCA", ncomps = 30, alignment.strength = 0.7, verbose = FALSE)
con$runEmbedding(method = "UMAP", embedding.name = "strength_0.7", min.visited.verts = 100, verbose = FALSE)
```

```{r}
#| fig-width: 6.5
#| fig-height: 5
con$plotGraph(embedding = "strength_0.7", color.by = "sample", mark.groups = FALSE, alpha = 0.1, show.legend = TRUE)
```

## Supervised alignment

Sometimes you know *which* factor the unwanted separation follows — here, **tissue** (BM vs CB). Instead of
turning up global strength, you can tell Conos to **down-weight edges within that factor**, so the alignment
works hardest *across* it. Build a per-cell tissue label from the sample names:

```{r}
tissue <- ifelse(grepl("BM", con$getDatasetPerCell()), "BM", "CB")
tissue <- setNames(factor(tissue), names(con$getDatasetPerCell()))
table(tissue)
```

`balancing.factor.per.cell` names the factor and `same.factor.downweight` (< 1) scales down within-factor
edges — so cross-tissue links carry relatively more weight:

```{r}
#| message: false
con$runGraph(k = 20, k.self = 5, space = "PCA", ncomps = 30,
             balancing.factor.per.cell = tissue, same.factor.downweight = 0.1,
             alignment.strength = 0.3, verbose = FALSE)
con$runEmbedding(method = "UMAP", embedding.name = "supervised", min.visited.verts = 100, verbose = FALSE)
```

```{r}
#| fig-width: 6.5
#| fig-height: 5
con$plotGraph(embedding = "supervised", color.by = "sample", mark.groups = FALSE, alpha = 0.1, show.legend = TRUE)
```

This targets the BM↔CB difference specifically, mixing across tissues without forcing every sample together
as hard as a high global `alignment.strength` would.

## Choosing a value

Reach for these only when the sample-mixing plot shows batch-like separation the default didn't remove:

* start from the **default** (no `alignment.strength`); raise it gradually (0.2–0.4 is usually plenty);
* prefer **supervised** alignment when you can name the nuisance factor (tissue, donor, batch) — it corrects
  that axis without over-mixing everything;
* after any change, re-run `runClustering()` and check markers (`plotMarkerDotPlot()`): if previously
  distinct populations have merged, you've gone too far.

The defaults are deliberately conservative — most panels need only a small nudge, if any.
