---
title: "Conos advanced: alignment spaces, memory, modalities, label transfer"
format:
  html:
    toc: true
    toc-depth: 3
    self-contained: true
  ipynb: default
jupyter: ir44
---

This tutorial assumes you have been through the [walkthrough](conos-walkthrough.ipynb) and covers the knobs
that matter when integration is harder: **which space** to align in, **how aggressively** to align,
**memory** on large panels, **what the samples actually share** (modalities/features), and **label-transfer**
choices. We reuse the same bone-marrow / cord-blood panel.

```{r}
#| message: false
#| warning: false
library(conos)
library(pagoda2)
library(conosPanel)

set.seed(1)
n.cores <- 4
panel <- lapply(conosPanel::panel, function(m) m[, sample(ncol(m), min(1000, ncol(m)))])  # subsample for speed
samples <- lapply(panel, function(cm) {
  rownames(cm) <- make.unique(rownames(cm))
  p <- Pagoda2$from(cm, n.cores = n.cores, min.cells.per.gene = 0, min.transcripts.per.cell = 0,
                    log.scale = TRUE, verbose = FALSE)
  p$run(steps = c("variance", "pca"), dependencies = "error", verbose = FALSE)   # just the reduction
  p
})
con <- Conos$new(samples, n.cores = n.cores)
```

## 1. Plan the integration first

`planIntegration()` is a read-only preliminary step. It polls each sample's molecular modalities (here just
RNA; for multimodal samples it would list ADT/ATAC facets too) and, per modality, reports how many features
are shared across the samples and how well they overlap — with a *usable / marginal / not-integrable*
verdict. This is where you catch problems like independently-called scATAC peaks that simply cannot be
reconciled without re-processing.

```{r}
plan <- con$planIntegration()
plan
```

The resolved default modality is recorded on the object and is what `runGraph()` aligns on unless you tell
it otherwise.

## 2. Choosing the alignment space

The `space` argument controls how each pair of samples is matched:

* **`"PCA"` (default)** — *reciprocal PCA*: each sample's PCA loadings define a shared gene basis both
  samples project into. Fast and robust; the right default for most panels.
* **`"CCA"`** — canonical correlation. Useful for **heterogeneous** integrations (very different
  compositions), but heavier and prone to **over-smoothing** (merging distinct populations) when little
  correction is actually needed.
* **`"CPCA"`** — common principal components; available, but in practice much slower without a clear gain
  over reciprocal PCA, so it is not the default.

```{r}
#| message: false
con$runGraph(k = 20, k.self = 5, space = "PCA", ncomps = 30, n.odgenes = 2000, verbose = FALSE)
con$runClustering(name = "leiden_pca", verbose = FALSE)

con$runGraph(k = 20, k.self = 5, space = "CCA", ncomps = 30, n.odgenes = 2000, verbose = FALSE)
con$runClustering(name = "leiden_cca", verbose = FALSE)

c(PCA = length(unique(con$clusters$leiden_pca$groups)),
  CCA = length(unique(con$clusters$leiden_cca$groups)))
```

CCA typically yields a more aggressively mixed graph (often fewer, larger clusters). Inspect both and pick
the one whose mixing matches your biological expectation rather than trusting a single default.

`alignment.strength` (and `k`/`k1`) tune how forcefully samples are pulled together — raise it for stronger
batch correction, lower it to preserve subtle structure.

## 3. Memory on large panels

The per-pair alignment rotations are O(samples²) and dominate memory on big panels. `pairs.storage` controls
them: `"keep"` (default) retains them in RAM for reuse, `"drop"` discards them after the graph is built
(a later `runGraph()` recomputes), and `"disk"` offloads them to a scratch file — freed from RAM but
restored on the next `runGraph()` so re-runs are cheap.

The disk path costs almost nothing in time (the rotations are written once after the graph is built), so on
large panels it is essentially free memory:

```{r}
#| message: false
t_keep <- system.time(con$runGraph(k = 20, k.self = 5, space = "PCA", ncomps = 30, n.odgenes = 2000,
                                   pairs.storage = "keep", verbose = FALSE))["elapsed"]
rot_mb  <- as.numeric(object.size(con$pairs[["PCA"]])) / 1e6     # rotation RAM held under "keep"

t_disk <- system.time(con$runGraph(k = 20, k.self = 5, space = "PCA", ncomps = 30, n.odgenes = 2000,
                                   pairs.storage = "disk", verbose = FALSE))["elapsed"]

c(keep_s = round(t_keep, 2), disk_s = round(t_disk, 2),
  rotation_RAM_freed_MB = round(rot_mb, 1), pairs_in_RAM_after_disk = length(con$pairs[["PCA"]]))
```

`runClustering()` / `runEmbedding()` work the same way regardless of `pairs.storage`. On a four-sample panel
the freed RAM is small, but it grows as the square of the number of samples, so `"disk"` is the right choice
for large atlases. (On lstar-backed pagoda2 samples the build path also streams expression in gene blocks,
keeping the per-sample footprint flat as well.)

```{r}
#| message: false
con$runClustering(verbose = FALSE)
```

## 4. Choosing a clustering resolution

`scanResolution()` runs the community detection over a range of resolutions and reports the number of
clusters (and modularity), helping you pick a value rather than guessing:

```{r}
#| fig-width: 6
#| fig-height: 4
res <- scanResolution(con, resolutions = c(0.5, 1, 1.5, 2), plot = TRUE, verbose = FALSE)
res
```

```{r}
#| message: false
con$runClustering(resolution = 1.5, name = "leiden_r15", verbose = FALSE)
```

## 5. Label transfer: diffusion vs solver

`propagateLabels()` offers two engines. On real panels they give near-identical results; the difference is
scaling:

* **`"diffusion"` (default)** — iterative graph diffusion. Scales to very large panels.
* **`"solver"`** — a Zhu–Ghahramani–Lafferty harmonic solve. Fast and slightly more accurate at typical
  sizes, but its linear solve does not scale much beyond ~20k cells.

```{r}
groups <- con$clusters$leiden_pca$groups
set.seed(1)
reference <- groups[sample(length(groups), round(0.30 * length(groups)))]
held.out  <- setdiff(names(groups), names(reference))

td <- con$propagateLabels(labels = reference, method = "diffusion", verbose = FALSE)
ts <- con$propagateLabels(labels = reference, method = "solver", solver = "Matrix")

c(diffusion = mean(td$labels[held.out] == as.character(groups[held.out])),
  solver    = mean(ts$labels[held.out] == as.character(groups[held.out])),
  agreement = mean(td$labels[held.out] == ts$labels[held.out]))
```

Use `diffusion` (the default) for very large atlases and `solver` when the panel is modest and you want the
extra speed/accuracy.

## 6. Multimodal panels (facets)

When samples carry more than one molecular modality (e.g. RNA + ADT in CITE-seq, or RNA + ATAC in multiome,
produced with pagoda2.1 *facets*), `planIntegration()` polls and assesses **each** modality separately, so
you can see which ones are actually integrable across the panel before committing. Today Conos aligns on a
single shared modality (the common default, usually RNA) — the recommended, robust path — and
modality-aware accessors (`getModalities()`, `getDefaultModality()`) expose the rest. Joint multimodal
*fusion* (weighting modalities per cell, à la WNN) is the next step on the roadmap.

## Recap

`planIntegration()` to see what's realistic → `runGraph(space=, alignment.strength=, pairs.storage=)` to
control the alignment and memory → `scanResolution()` + `runClustering(resolution=)` to tune granularity →
`propagateLabels(method=)` to transfer annotations. Start from the [walkthrough](conos-walkthrough.ipynb)
defaults and reach for these only when a particular integration needs them.
